---
title: "Red Wine Physicochemical Properties and Quality"
author: "Oliver Koch"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
   html_document:
      # theme of html document
      # theme of code highlight
      # table of contents
      theme          : cosmo     # "default", "cerulean", "journal",
                                 # "flatly", "readable", "spacelab",
                                 # "united", "cosmo", "lumen", "paper",
                                 # "sandstone", "simplex", "yeti"
      highlight      : tango     # "default", "tango", "pygments",
                                 # "kate", "monochrome", "espresso",
                                 # "zenburn", "haddock", "textmate"
      toc            : true
      toc_depth      : 3
      toc_float      : true      # get table of content
      code_folding   : hide
      number_sections: yes
      fig_caption    : yes
      df_print       : paged
---

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using in your analysis in this code
# chunk.

# Notice that the parameter "echo" was set to FALSE for this code chunk. This
# prevents the code from displaying in the knitted HTML output. You should set
# echo=FALSE for all code chunks in your file, unless it makes sense for your
# report to show the code that generated a particular plot.

# The other parameters for "message" and "warning" should also be set to FALSE
# for other code chunks once you have verified that each plot comes out as you
# want it to. This will clean up the flow of your report.

library(readr)
library(tidyr)
library(knitr)
library(ggplot2)
library(dplyr)
library(GGally)
library(tibble)

# knitr: suppress code/messages/warnings
#        set default plot options and center them
knitr::opts_chunk$set(fig.width=9, fig.height=5, fig.path='Figs/', 
                      fig.align='center', tidy=TRUE, echo=FALSE, warning=FALSE, 
                      message=FALSE)

```

```{r echo=FALSE, message=FALSE, warning=FALSE, Load_the_Data}
# Load the Data

# By default, read_csv reads numerical data as float. I want the quality 
# column as an integer because that later conversion to a categorical variable 
# easier.

wine <- read_csv("wineQualityReds.csv", 
                      col_types = cols(quality = col_integer()))

# The first column of the csv file contains a row index without column name.
# R automatically assigns this column a name 'X1' --- I rename it to 'ID'.

wine <- reshape::rename(wine, c(X1 = "ID"))
```


## Introduction to the Dataset

The dataset is related to red variants of the Portuguese "Vinho 
Verde" wine. It contains physical and chemical properties from objective 
measurements and a sensory assessment of the quality of the wine (median of at 
least 3 evaluations made by wine experts).

The dataset is available at this URL:
https://s3.amazonaws.com/udacity-hosted-downloads/ud651/wineQualityReds.csv

The details are described in 
_P. Cortez, A. Cerdeira, F. Almeida, T. Matos and J. Reis. 
Modeling wine preferences by data mining from physicochemical properties.
In Decision Support Systems, Elsevier, 47(4):547-553. ISSN: 0167-9236_, 
available at http://www3.dsi.uminho.pt/pcortez/wine5.pdf


### Attribute information:

Input variables (based on physicochemical tests):

1. Fixed acidity (tartaric acid - g / dm^3): most acids involved with wine or 
   fixed or nonvolatile (do not evaporate readily)
2. Volatile acidity (acetic acid - g / dm^3): the amount of acetic acid in wine, 
   which at too high of levels can lead to an unpleasant, vinegar taste
3. Citric acid (g / dm^3): found in small quantities, citric acid can add 
   'freshness' and flavor to wines
4. Residual sugar (g / dm^3): the amount of sugar remaining after fermentation 
   stops, it's rare to find wines with less than 1 gram/liter and wines with 
   greater than 45 grams/liter are considered sweet
5. Chlorides (sodium chloride - g / dm^3): the amount of salt in the wine
6. Free sulfur dioxide (mg / dm^3): the free form of SO2 exists in equilibrium 
   between molecular SO2 (as a dissolved gas) and bisulfite ion; it prevents 
   microbial growth and the oxidation of wine
7. Total sulfur dioxide (mg / dm^3): amount of free and bound forms of S02; in 
   low concentrations, SO2 is mostly undetectable in wine, but at free SO2 
   concentrations over 50 ppm, SO2 becomes evident in the nose and taste of wine
8. Density (g / cm^3): the density of water is close to that of water depending 
   on the percent alcohol and sugar content
9. pH: describes how acidic or basic a wine is on a scale from 0 (very acidic) 
   to 14 (very basic); most wines are between 3-4 on the pH scale
10. Sulphates (potassium sulphate - g / dm3): sulphates: a wine additive which 
    can contribute to sulfur dioxide gas (S02) levels, wich acts as an 
    antimicrobial and antioxidant
11. Alcohol (% by volume): the percent alcohol content of the wine

Output variable (based on sensory data): 

12. Quality. Score between 0 (very bad) and 10 (very good)


# Univariate Plots Section

An array of histograms and an array of boxplots provide a first overview of the 
ranges of values and the distribution of data:

```{r echo=FALSE, message=FALSE, warning=FALSE, histo_1}

# create a grid of histograms. 

# First, reshape dataframe into a long list of key/value pairs (using gather), 
# thereby omitting the ID variable. The original column names go into the 
# 'wine.att' column, values into the 'measurement' column.

# 'facet_wrap' then creates a faceted view - one histogram per attribute. 

wine %>%
gather(key="wine.att", value="measurement", -ID) %>%
ggplot() +
 geom_histogram(mapping=aes(measurement), bins=50) +
 facet_wrap(vars(wine.att), scales='free')
```
```{r echo=FALSE, message=FALSE, warning=FALSE, box_1}

# making a grid of boxplots.

# First, The select function selects all columns except ID and quality.

# Second, reshape dataframe into a long list of key/value pairs (using gather).
# as for the histograms above the original column names go into the 'wine.att' 
# column, values into the 'measurement' column.
# 'facet_wrap' then creates a faceted view - one boxplot per attribute. 

wine %>%
select(-c(ID, quality)) %>%
gather(key="wine.att", value="measurement") %>%
ggplot(mapping=aes(measurement)) +
 geom_boxplot() +
 facet_wrap(vars(wine.att), scales='free')
```

At a first glance, pH, density and quality have symmetric, normal-like 
distributions. The other variables are right skewed.

Plotting the histograms again with a logarithmic scale provides a better view on 
extreme values:

```{r echo=FALSE, message=FALSE, warning=FALSE, histo_2}

# same procedure as in histo_1, except for the logarithmic x axis.

wine %>%
  gather(key="wine.att", value="measurement", -ID) %>%
  ggplot() +
   geom_histogram(mapping=aes(measurement), bins=50) +
   scale_x_log10() +
   facet_wrap(vars(wine.att), scales='free')
```

In order to better visualize the variables with long-tailed distributions, I 
create a filtered dataset (with manually set limits per variable) and plot the 
histograms again:

```{r echo=FALSE, message=FALSE, warning=FALSE, histo_3}

# Create a dataframe without extreme values. Criteria were chosen manually.
# A more systematic approach would be to use quantiles.

wine.typical <- filter(wine, alcohol < 13 & 
                              chlorides < 0.15 & 
                              citric.acid > 0.01 &
                              citric.acid < 0.8 & 
                              fixed.acidity < 12 &
                              free.sulfur.dioxide < 40 &
                              residual.sugar < 3.5 &
                              sulphates < 1 &
                              total.sulfur.dioxide < 100 &
                              volatile.acidity < 1
                           ) 

# A grid of histograms as in histo_2.
# Additional filter: only measurements > 0 because of logarithmic x axis.

wine.typical %>%
gather(key="wine.att", value="measurement", -ID) %>%
filter(measurement > 0) %>%                      # because of log
ggplot() +
 geom_histogram(mapping=aes(measurement), bins=20) +
 scale_x_log10() +
 facet_wrap(vars(wine.att), scales='free')
```

Looking at quality:

```{r echo=FALSE, message=FALSE, warning=FALSE, histo_4}

# a simple histogram of wine quality.

wine %>%
ggplot() +
 geom_histogram(mapping=aes(quality)) 
```

```{r echo=FALSE, message=FALSE, warning=FALSE, table_1}
# show counts per quality level
table(wine$quality)
```

```{r echo=FALSE, message=FALSE, warning=FALSE, summary_1}
summary(wine$quality)
```

Quality can take values form 0 to 10. The wines present in the dataset only span 
a range from 3 to 8. There are no _very good_ or _very bad_ wines; the majority 
has a quality of 5 or 6.

Introducing a new categorical variable, reducing quality to three grades:

- 0-4: bad
- 5-6: medium
- 7-10: good

```{r echo=FALSE, message=FALSE, warning=FALSE, categories}

# create a vector with grades
quality_grades <- ifelse(wine$quality <= 4, "bad", 
                         ifelse(wine$quality >= 7, "good", "medium"))

# make an ordered categorical variable
wine$grade <- ordered(quality_grades, levels = c("bad", "medium", "good"))

# show the distribution
table(wine$grade)
```

There are only very few bad wines.

```{r echo=FALSE, message=FALSE, warning=FALSE, bar_1}

# for categorical variables, use bar plot instead of histograms
ggplot(wine) +
 geom_bar(mapping=aes(grade)) 
```

Quality is distributed very unevenly - the majority of wines is in the "medium" 
group, and there are only very few bad wines.


# Univariate Analysis

Summary statistics of data:

```{r echo=FALSE, message=FALSE, warning=FALSE, analysis_1}

# make a summary, omit ID column
wine %>%
select(-ID) %>%
summary()
```

### What is the structure of your dataset?

```{r echo=FALSE, message=FALSE, warning=FALSE, dim_1}

# dim shows dataframe size (num rows, num cols). 14 columns including ID and the 
# 'grade' category, but only 12 original variables.´

dim(wine)
```

The dataset has 12 variables and 1599 observations. All variables except quality
are real numbers. Quality is an integer variable.
There are no missing values. There are no obviously incorrect values (like 
negative values.)


### What is/are the main feature(s) of interest in your dataset?

The wine quality is the main feature of interest. As suggested in the datasets
documentation, it was created the purpose of developing regression models that
can predict the wine quality from physical & chemical measurement results.


### What other features in the dataset do you think will help support your investigation into your feature(s) of interest?

To me, the most interesting question is which variables have the strongest 
influence on wine quality. Potentially, all variables may have an influence.


### Did you create any new variables from existing variables in the dataset?

I created a categorical variable 'grade' from quality. 


### Of the features you investigated, were there any unusual distributions? Did you perform any operations on the data to tidy, adjust, or change the form of the data? If so, why did you do this?

Alcohol and citric acid content have highly asymmetrical, almost triangular 
distributions. 

Quality has a bell-shaped distribution, there are few good ones and even 
fewer bad ones. 


# Bivariate Plots Section

A pair plot (scatterplot matrix) provides a first overview of dependencies and 
potential correlations between variables:

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.width=9, fig.height=8, spm1}

# make a scatterplot matrix.

# ignore ID and grade columns. 

# the 'upper = ...' parameter in ggpairs results in display of correlation
# values instead of default mirrored and redundant scatter plots.

# The remainder of the options makes sure that all information in this complex
# plot is readable.

wine %>%
   select(-c(ID, grade)) %>% 
   ggpairs(upper = list(continuous = wrap("cor", size=2.7))) +
   theme_grey(base_size=5) +
   theme(text = element_text(size=8),         
         axis.text.x = element_text(angle=45, vjust=1, hjust=1, size=7),
         panel.grid.minor = element_blank(), 
         panel.grid.major = element_blank())  # rotate and resize axis labels
                                              # and remove grid lines
```

Again a pair plot, this time using the categorical quality value:

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.width=9, fig.height=8, spm2}

# Another scatterplot matrix - this time using the categorical quality grade
# instead of the numerical value. 

# ignore ID and quality columns
# for further comments see 'spm1' code chunk.

wine %>%
  select(-c(ID, quality)) %>%
  ggpairs(upper = list(continuous = wrap("cor", size=1.8)), 
          mapping = aes(color = grade)) +
  theme_grey(base_size=5) +
  theme(text = element_text(size=8), 
        axis.text.x = element_text(angle=45, vjust=1, hjust=1, size=7),
        panel.grid.minor = element_blank(), 
        panel.grid.major = element_blank())  # rotate and resize axis labels
                                              # and remove grid lines   
```

For the main variable of interest (quality), the highest correlations 
are with volatile acidity, alcohol, and sulphates. 

High values of correlation also occur in these groups: 

- volatile acidity, fixed.acidity, citric acid, density, pH
- density, alcohol, pH, fixed.acidity, citric.acid, residual.sugar


### Which variables influence quality?

The following plots highlight patterns between the measured variables and the 
wine quality:

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.height=7, influence_quality}

# Grid of scatterplots focusing on quality

# Reshape dataframe into a long form with columns ID, quality, grade, wine.att
# and measurement.

# 'facet_wrap' then creates a faceted view - one plot per attribute. 
# jitter is used on (integer) y axis only, to give a better impression of the
# density of observations.
# Jitter on y only, using 0.35 to indicate visually that the data are integer / 
# categorical.

# geom_smooth is used to indicate trends.

wine %>%
gather(key="wine.att", value="measurement", -c(ID, quality, grade)) %>%
ggplot(aes(x=measurement, y=quality)) +
 geom_jitter(alpha=0.2, height=0.35, width=0) +
 geom_smooth() +
 facet_wrap(vars(wine.att), scales='free')
```

- alcohol, density, total.sulfur.dioxide, volatile acidity
- free sulfur dioxide and residual sugar do not have an influence. 
- nonlinear relationships (which are not captured by linear correlation measures 
  or linear models): citric acid, chlorides, sulphates


### Which variables influence the pH value?

```{r echo=FALSE, message=FALSE, warning=FALSE, influence_ph}

# Grid of scatterplots focusing on pH value

# First, select columns related to pH and acids only.
# Then, grid of scatter plots plus smoothing line, similar to the plot above.

wine %>%
select(-c(residual.sugar, chlorides, free.sulfur.dioxide, total.sulfur.dioxide, 
          density, sulphates, alcohol, quality, grade)) %>%
gather(key="wine.att", value="measurement", -c(ID, pH)) %>%
ggplot(aes(x=measurement, y=pH)) +
 geom_point(alpha=0.1) +
 geom_smooth() +
 facet_wrap(vars(wine.att), scales='free')
```

correlation between pH and fixed.acidity:

```{r echo=FALSE, message=FALSE, warning=FALSE}
cor(wine$fixed.acidity, wine$pH)
```

correlation between pH and volatile acidity:

```{r echo=FALSE, message=FALSE, warning=FALSE}
cor(wine$volatile.acidity, wine$pH)
```

correlation between pH and citric acid:

```{r echo=FALSE, message=FALSE, warning=FALSE}
cor(wine$citric.acid, wine$pH)
```

Conclusion:

- medium correlation between pH and citric.acid and fixed.acidity
- no clear correlation between pH and volatile.acidity



# Bivariate Analysis

### Talk about some of the relationships you observed in this part of the investigation. How did the feature(s) of interest vary with other features in the dataset?

The alcohol content has the strongest influence on wine quality. The variables
with the next highest correlation coefficients are sulphates and volatile 
acidity. 


### Did you observe any interesting relationships between the other features (not the main feature(s) of interest)?

Relationships exist 

- among acidity related variables (pH, citric acid, fixed acidity, volatile 
  acidity). 
- between density and almost all other variables (because the density of the 
  wine depends on the concentration of the ingredients). 
  Most prominent influences from alcohol and residual sugar.


### What was the strongest relationship you found?

The strongest (linear) relationship is between total sulfur dioxide and free 
sulfur dioxide.


# Multivariate Plots Section

### Explore relationship between quality, alcohol, volatile acidity, sulphates

```{r echo=FALSE, message=FALSE, warning=FALSE, Multivariate_Plots}

# simple xy plot, using the quality grade as color.

wine %>%
  ggplot() + 
   geom_point(mapping=aes(x=alcohol, y=volatile.acidity, color=grade), 
              alpha=0.9) 
```

```{r echo=FALSE, message=FALSE, warning=FALSE}

# simple xy plot, using the quality grade as color.

wine %>%
  ggplot() + 
   geom_point(mapping=aes(x=alcohol, y=sulphates, color=grade), 
              alpha=0.9) 
```


```{r echo=FALSE, message=FALSE, warning=FALSE}

# simple xy plot, using the quality grade as color.

wine %>%
  ggplot() + 
   geom_point(mapping=aes(x=volatile.acidity, y=sulphates, color=grade), 
              alpha=0.8) 

```

In the plots above, bad and good wines are separated - if there were just good 
ones and bad ones, the three variables could be used to classify wines.
However, the majority of wines are in the medium group, and they cover the 
whole range of values, making a classification (with the given variables and 
linear decision boundaries) difficult.


# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the investigation. Were there features that strengthened each other in terms of looking at your feature(s) of interest?

In the plots above, bad and good wines are separated - if there were just good 
ones and bad ones, the three variables could be used to classify wines.
However, the majority of wines are in the medium group, and they cover the 
whole range of values, making a classification (with the given variables and 
linear decision boundaries) difficult.

------

# Final Plots and Summary


### Plot One

```{r echo=FALSE, message=FALSE, warning=FALSE, Plot_One}

ggplot(data = wine) +
   geom_bar(aes(quality)) +
   labs(x = "Quality", title = "Distribution of wine quality") # custom labels
```

### Description One

On a possible range from 0 to 10, all wines fall 
between 3 and 8, and the majority has quality ratings of 5 or 6.


### Plot Two

```{r echo=FALSE, message=FALSE, warning=FALSE, Plot_Two}

# like spm2, but with a reduced set of variables and custom labels

wine %>%
  select(c(fixed.acidity, volatile.acidity, citric.acid, sulphates, alcohol, 
           grade)) %>%
  ggpairs(upper = list(continuous = wrap("cor", size=1.8)), 
          mapping = aes(color = grade)) +
  theme_grey(base_size=5) +
  theme(text = element_text(size=8), 
        axis.text.x = element_text(angle=45, vjust=1, hjust=1, size=7),
        panel.grid.minor = element_blank(), 
        panel.grid.major = element_blank()) + # rotate and resize axis labels
                                              # and remove grid lines   
        labs(title = "Scatterplot matrix for selected variables")   # custom labels
```

### Description Two

The scatterplot matrix shows those variable where distributions differ among  
wine "grades". This subset of variables could be a starting point for building 
classification or regression models.


### Plot Three

```{r echo=FALSE, message=FALSE, warning=FALSE, Plot_Three}

# as in influence_quality, but reduced set of variables, custom labels

wine %>%
gather(key="wine.att", value="Attribute", 
       -c(ID, quality, grade, fixed.acidity, free.sulfur.dioxide, residual.sugar, 
          pH, density)) %>%
ggplot(aes(x=Attribute, y=quality)) +
 geom_jitter(alpha=0.5, height = 0.35, width = 0) +
 geom_smooth() +
 facet_wrap(vars(wine.att), scales='free') +
 labs(title = "Linear and nonlinear relationships between quality and other attributes", 
      y = "Quality")

```

### Description Three

This plot shows that there are nonlinear relationships between quality and some
of the other variables, e.g. citric acid. A good classification or regression 
model would have to capture these nonlinear relationships.

------

# Reflection

The dataset can be used to predict the wine quality from physical/chemical 
measurements. 

Difficulties with this dataset: 

- the majority of data lies in the medium range; there are relatively few 'good' 
  or 'bad' examples. There are no samples for extreme cases (quality below three 
  or above 8).
- There are nonlinear relationships between some variables (e.g. citric acid and 
  quality). Linear correlation measures do not capture these. 

Successes:

- Alcohol content is a good indicator for wine quality
- There are some correlations between the variables (e.g. fixed acidity and pH). 
  These could be used to eliminate some of the variables and reduce the number 
  of dimensions of the problem.

Enriching the analysis:

- More data for extreme cases (good / bad wines)
- use nonlinear models
- use dimensionality reduction



