TITLE by YOUR_NAME_HERE
========================================================

> **Tip**: You will see quoted sections like this throughout the template to
help you construct your report. Make sure that you remove these notes before
you finish and submit your project!

> **Tip**: One of the requirements of this project is that your code follows
good formatting techniques, including limiting your lines to 80 characters or
less. If you're using RStudio, go into Preferences \> Code \> Display to set up
a margin line to help you keep track of this guideline!

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using in your analysis in this code
# chunk.

# Notice that the parameter "echo" was set to FALSE for this code chunk. This
# prevents the code from displaying in the knitted HTML output. You should set
# echo=FALSE for all code chunks in your file, unless it makes sense for your
# report to show the code that generated a particular plot.

# The other parameters for "message" and "warning" should also be set to FALSE
# for other code chunks once you have verified that each plot comes out as you
# want it to. This will clean up the flow of your report.

library(tidyverse)
library(GGally)

```

```{r echo=FALSE, warning=FALSE, Load_the_Data}
# Load the Data

# the first column of the csv file contains a row index without column name.
# R automatically assigns this column a name 'X1'

wine <- read_csv("wineQualityReds.csv", 
                      col_types = cols(quality = col_integer()))
wine <- reshape::rename(wine, c(X1 = "ID"))
```

> **Tip**: Before you create any plots, it is a good idea to provide a short
introduction into the dataset that you are planning to explore. Replace this
quoted text with that general information!


#### Citation Request:

This dataset is public available for research. The details are described in 
[Cortez et al., 2009]. 
Please include this citation if you plan to use this database:

P. Cortez, A. Cerdeira, F. Almeida, T. Matos and J. Reis. 
Modeling wine preferences by data mining from physicochemical properties.
In Decision Support Systems, Elsevier, 47(4):547-553. ISSN: 0167-9236.

Available at: 

- [@Elsevier] http://dx.doi.org/10.1016/j.dss.2009.05.016
- [Pre-press (pdf)] http://www3.dsi.uminho.pt/pcortez/winequality09.pdf (dead link)
- [bib] http://www3.dsi.uminho.pt/pcortez/dss09.bib

here: http://www3.dsi.uminho.pt/pcortez/wine5.pdf

#### Title: Wine Quality 

#### Sources

Created by: Paulo Cortez (Univ. Minho), Antonio Cerdeira, Fernando Almeida, 
Telmo Matos and Jose Reis (CVRVV) @ 2009


In the above reference, two datasets were created, using red and white wine 
samples. The inputs include objective tests (e.g. PH values) and the output is 
based on sensory data (median of at least 3 evaluations made by wine experts). 
Each expert graded the wine quality between 0 (very bad) and 10 (very 
excellent). Several data mining methods were applied to model these datasets 
under a regression approach. The support vector machine model achieved the
best results. Several metrics were computed: MAD, confusion matrix for a fixed 
error tolerance (T), etc. Also, we plot the relative importances of the input 
variables (as measured by a sensitivity analysis procedure).
 
#### Relevant Information:

The two datasets are related to red and white variants of the Portuguese "Vinho 
Verde" wine. For more details, consult: http://www.vinhoverde.pt/en/ or the 
reference [Cortez et al., 2009]. Due to privacy and logistic issues, only 
physicochemical (inputs) and sensory (the output) variables are available (e.g. 
there is no data about grape types, wine brand, wine selling price, etc.).

These datasets can be viewed as classification or regression tasks.
The classes are ordered and not balanced (e.g. there are munch more normal wines 
than excellent or poor ones). Outlier detection algorithms could be used to 
detect the few excellent or poor wines. Also, we are not sure if all input 
variables are relevant. So it could be interesting to test feature selection 
methods. 

#### Number of Instances: red wine - 1599; white wine - 4898. 

#### Number of Attributes: 11 + output attribute

Note: several of the attributes may be correlated, thus it makes sense to apply 
some sort of feature selection.

#### Attribute information:

For more information, read [Cortez et al., 2009].

Input variables (based on physicochemical tests):

1. fixed acidity (tartaric acid - g / dm^3): most acids involved with wine or 
   fixed or nonvolatile (do not evaporate readily)
2. volatile acidity (acetic acid - g / dm^3): the amount of acetic acid in wine, 
   which at too high of levels can lead to an unpleasant, vinegar taste
3. citric acid (g / dm^3): found in small quantities, citric acid can add 
   'freshness' and flavor to wines
4. residual sugar (g / dm^3): the amount of sugar remaining after fermentation 
   stops, it's rare to find wines with less than 1 gram/liter and wines with 
   greater than 45 grams/liter are considered sweet
5. chlorides (sodium chloride - g / dm^3): the amount of salt in the wine
6. free sulfur dioxide (mg / dm^3): the free form of SO2 exists in equilibrium 
   between molecular SO2 (as a dissolved gas) and bisulfite ion; it prevents 
   microbial growth and the oxidation of wine
7. total sulfur dioxide (mg / dm^3): amount of free and bound forms of S02; in 
   low concentrations, SO2 is mostly undetectable in wine, but at free SO2 
   concentrations over 50 ppm, SO2 becomes evident in the nose and taste of wine
8. density (g / cm^3): the density of water is close to that of water depending 
   on the percent alcohol and sugar content
9. pH: describes how acidic or basic a wine is on a scale from 0 (very acidic) 
   to 14 (very basic); most wines are between 3-4 on the pH scale
10. sulphates (potassium sulphate - g / dm3): sulphates: a wine additive which 
    can contribute to sulfur dioxide gas (S02) levels, wich acts as an 
    antimicrobial and antioxidant
11. alcohol (% by volume): the percent alcohol content of the wine

Output variable (based on sensory data): 

12. quality (score between 0 and 10)

#### Missing Attribute Values

None


# Univariate Plots Section

> **Tip**: In this section, you should perform some preliminary exploration of
your dataset. Run some summaries of the data and create univariate plots to
understand the structure of the individual variables in your dataset. Don't
forget to add a comment after each plot or closely-related group of plots!
There should be multiple code chunks and text sections; the first one below is
just to help you get started.



```{r echo=FALSE, warning=FALSE, Univariate_Plots}

wine %>%
# sample_n(100) %>%
gather(key="wine.att", value="measurement", -ID) %>%
ggplot() +
 geom_histogram(mapping=aes(measurement), bins=30) +
 facet_wrap(vars(wine.att), scales='free')
  
```

- residual sugar: a peak 
- sulphates: bimodal?
- most distributions are right skewed
- log plot or remove outliers?

- cut-off values: manual or by percentile?

```{r echo=FALSE, warning=FALSE, Univariate_Plots}

wine %>%
select(-c(ID, quality)) %>%
gather(key="wine.att", value="measurement") %>%
ggplot(mapping=aes(measurement)) +
 geom_boxplot() +
 facet_wrap(vars(wine.att), scales='free')
  
```
```{r}
# dataset without extreme values
wine %>%
# sample_n(100) %>%
# gather(key="wine.att", value="measurement", -ID) %>%
summarise()
```



```{r}
# dataset without extreme values
wine.typical <- filter(wine, alcohol < 13 & 
                              chlorides < 0.15 & 
                              citric.acid < 0.8 & 
                              fixed.acidity < 12 &
                              free.sulfur.dioxide < 40 &
                              residual.sugar < 3.5 &
                              sulphates < 1 &
                              total.sulfur.dioxide < 100 &
                              volatile.acidity < 1
                           ) 

wine.typical %>%
gather(key="wine.att", value="measurement", -ID) %>%
ggplot() +
 geom_histogram(mapping=aes(measurement), bins=20) +
 facet_wrap(vars(wine.att), scales='free')
```

```{r}

wine.typical %>%
  # sample_n(100) %>%
gather(key="wine.att", value="measurement", -ID)%>%
filter(measurement > 0) %>% # cecause of log
ggplot() +
  geom_histogram(mapping=aes(measurement),  bins=30) +
  scale_x_log10() +
  facet_wrap(vars(wine.att), scales='free')
```

```{r}


# log log auch ok ?!?!
wine %>%
select(c(ID, alcohol, citric.acid)) %>%
gather(key="wine.att", value="measurement", -ID)%>%
filter(measurement > 0) %>% # cecause of log
ggplot() +
  geom_histogram(mapping=aes(measurement),  bins=40) +
  scale_x_log10() +
  #scale_y_log10() +
  facet_wrap(vars(wine.att), scales='free')
```

- x-log: bimodale verteilung bei residual sugar




> **Tip**: Make sure that you leave a blank line between the start / end of
each code block and the end / start of your Markdown text so that it is
formatted nicely in the knitted text. Note as well that text on consecutive
lines is treated as a single space. Make sure you have a blank line between
your paragraphs so that they too are formatted for easy readability.

# Univariate Analysis

> **Tip**: Now that you've completed your univariate explorations, it's time to
reflect on and summarize what you've found. Use the questions below to help you
gather your observations and add your own if you have other thoughts!

```{r}

wine %>%
select(-ID) %>%
summary()
  
```

### What is the structure of your dataset?

- num columns, num rows

### What is/are the main feature(s) of interest in your dataset?

- wine quality

### What other features in the dataset do you think will help support your \
investigation into your feature(s) of interest?

- how do other variables influence wine quality?
- which are important, which not?
- can we distinguish really good ones from really bad ones?

### Did you create any new variables from existing variables in the dataset?

- could do residual content: total - sum_i rho_i pct_i
- factor: bad / medium / good

### Of the features you investigated, were there any unusual distributions? \
Did you perform any operations on the data to tidy, adjust, or change the form \
of the data? If so, why did you do this?

- removed some extreme value
- mostly normally distributed


# Bivariate Plots Section

> **Tip**: Based on what you saw in the univariate plots, what relationships
between variables might be interesting to look at in this section? Don't limit
yourself to relationships between a main output feature and one of the
supporting variables. Try to look at relationships between supporting variables
as well.

```{r echo=FALSE, Bivariate_Plots}

wine.typical %>%
   select(-ID) %>%
   ggpairs()
```

```{r echo=FALSE, Bivariate_Plots}

wine.typical %>%
#sample_n(300) %>%
gather(key="wine.att", value="measurement", -c(ID, quality))%>%
ggplot(aes(x=measurement, y=quality)) +
 #geom_jitter(width=0, height=0.5) +
 geom_point(alpha=0.1) +
 # scale_x_sqrt() +
 #scale_x_log10() +
 geom_smooth() +
 #geom_quantile() +
 facet_wrap(vars(wine.att), scales='free')
```

#### related: pH, fixed.acidity, volatile.acidity, citric.acid

```{r echo=FALSE, acidity}
wine.typical %>%
select(-c(residual.sugar, chlorides, free.sulfur.dioxide, total.sulfur.dioxide, density, sulphates, alcohol, quality)) %>%
gather(key="wine.att", value="measurement", -c(ID, pH)) %>%
ggplot(aes(x=measurement, y=pH)) +
 #geom_jitter(width=0, height=0.5) +
 geom_point(alpha=0.1) +
 # scale_x_sqrt() +
 #scale_x_log10() +
 geom_smooth() +
 #geom_quantile() +
 facet_wrap(vars(wine.att), scales='free')
```

- dependency between pH and fixed.acidity
- no clear dependency on volatile.acidity

#### related: free.sulfur.dioxide, total.sulfur.dioxide

```{r echo=FALSE, so2}

ggplot(data=wine, aes(x=free.sulfur.dioxide, y=total.sulfur.dioxide)) +
   geom_point() +
   geom_abline(slope=1, intercept=0)

```
```{r echo=FALSE, so2}
bound.sulfur.dioxide <- wine$total.sulfur.dioxide - wine$free.sulfur.dioxide
cor(wine$total.sulfur.dioxide, wine$free.sulfur.dioxide)
cor(bound.sulfur.dioxide, wine$free.sulfur.dioxide)
```


```{r echo=FALSE, so2}

wine.typical %>%
   mutate(bound.sulfur.dioxide = total.sulfur.dioxide - free.sulfur.dioxide) %>%
   ggplot(aes(x = bound.sulfur.dioxide, y=free.sulfur.dioxide)) +
   geom_point() +
   geom_smooth()

wine.typical %>%
   mutate(bound.sulfur.dioxide = total.sulfur.dioxide - free.sulfur.dioxide) %>%
   ggplot(aes(x = bound.sulfur.dioxide, y=free.sulfur.dioxide)) +
   geom_point() +
   #scale_x_log10() +
   scale_y_log10() +
   geom_smooth()

```

- all points above line
- could introduce a new variable: bound = total - free

Intro says that for free SO2 > 50 ppm quality suffers?
Units are mg/dm^3 -> that is 

```{r echo=FALSE, so2}


wine %>%
#select(c(free.sulfur.dioxide, quality)) %>%
#gather(key="wine.att", value="measurement") %>%
ggplot(mapping=aes(free.sulfur.dioxide)) +
 geom_boxplot() +
 facet_grid(rows = vars(quality), scales='free')
  
```
```{r echo=FALSE, so2}

wine %>%
   filter(free.sulfur.dioxide >= 50) %>%
   select(quality) %>%
   table
```

- wines with free.sulfur.dioxide > 50 are in medium to good range
- wines of quality 5 have the highest median free.sulfur.dioxide
- with increasing quality, median free.sulfur.dioxide decreases
- with decreasing quality, median free.sulfur.dioxide decreases
- statement from dataset description cannot be observed in this dataset.


```{r echo=FALSE, sulfates}


wine %>%
#select(c(free.sulfur.dioxide, quality)) %>%
#gather(key="wine.att", value="measurement") %>%
ggplot(mapping=aes(sulphates)) +
 geom_boxplot() +
 facet_grid(rows = vars(quality), scales='free')
  
```
- good wines tend to have higher sulphate content (median 0.7), but not too high
- wines with sulphate content > 1.4 don't make it into the upper range

```{r echo=FALSE, so2}
wine %>%
   select(c(sulphates, quality)) %>%
   filter(quality >= 7) %>%
   summary()
```

```{r echo=FALSE, so2}
wine %>%
   select(c(sulphates, quality)) %>%
   filter(quality <= 4) %>%
   summary()
```

#### related: density, alcohol, sugar


```{r echo=FALSE, so2}
ggplot(data=wine, aes(x=residual.sugar, y=density)) + 
   geom_point() +
   #scale_x_log10() +
   coord_cartesian(xlim = c(0, 4)) +
   geom_smooth() 

ggplot(data=wine, aes(x=alcohol, y=density)) + 
   geom_point() +
   geom_smooth() 
```

build a linear model: density depending on 


# Bivariate Analysis

> **Tip**: As before, summarize what you found in your bivariate explorations
here. Use the questions below to guide your discussion.

### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?

### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?

### What was the strongest relationship you found?

- alcohol / sugar / acidity -> quality
- alcohol / sugar -> density
- there is some redundancy here: density depends on alcohol, sugar and other
  stuff

# Multivariate Plots Section

> **Tip**: Now it's time to put everything together. Based on what you found in
the bivariate plots section, create a few multivariate plots to investigate
more complex interactions between variables. Make sure that the plots that you
create here are justified by the plots you explored in the previous section. If
you plan on creating any mathematical models, this is the section where you
will do that.

```{r echo=FALSE, Multivariate_Plots}

```

#### sulfates / alcohol / quality
```{r echo=FALSE, three vars}
#saq <- select(wine.filtered, c(volatile.acidity, alcohol, quality))

#ggplot(data=saq) +
#   geom_bin2d(aes(x=volatile.acidity, y=alcohol))
   #geom_point() +
   #geom_contour(aes(z=quality))

qlevel <- ordered(wine$quality, levels = 0:10)
wine %>%
  add_column(qlevel = qlevel) %>%
  ggplot() + 
   # geom_point(mapping=aes(x=alcohol, y=volatile.acidity, color=qlevel), alpha=0.9) +
   geom_jitter(mapping=aes(x=alcohol, y=volatile.acidity, color=qlevel), alpha=0.9, width=0.005, height=0) +
   #geom_contour(aes(x=alcohol, y=volatile.acidity, z=qlevel), )
   #scale_x_log10() +
   scale_y_sqrt()


```

- good wines tend to have high alcohol and low volatile acidity

```{r echo=FALSE, Multivariate_Plots}
qlevel <- ordered(wine$quality, levels = 0:10)
wine %>%
  add_column(qlevel = qlevel) %>%
  ggplot() + 
   # geom_point(mapping=aes(x=alcohol, y=volatile.acidity, color=qlevel), alpha=0.9) +
   geom_jitter(mapping=aes(x=alcohol, y=sulphates, color=qlevel), alpha=0.9, width=0.005, height=0) 
   #geom_contour(aes(x=alcohol, y=volatile.acidity, z=qlevel), )
   #scale_x_log10() +
   scale_y_log10()

```



```{r echo=FALSE, three vars}
wine %>%
   select(c(volatile.acidity, alcohol, quality)) %>%
   ggplot() +
   geom_point(aes(x=volatile.acidity, y=alcohol))

```


```

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?

### Were there any interesting or surprising interactions between features?

### OPTIONAL: Did you create any models with your dataset? Discuss the \
strengths and limitations of your model.





------

# Final Plots and Summary

> **Tip**: You've done a lot of exploration and have built up an understanding
of the structure of and relationships between the variables in your dataset.
Here, you will select three plots from all of your previous exploration to
present here as a summary of some of your most interesting findings. Make sure
that you have refined your selected plots for good titling, axis labels (with
units), and good aesthetic choices (e.g. color, transparency). After each plot,
make sure you justify why you chose each plot by describing what it shows.

### Plot One
```{r echo=FALSE, Plot_One}

```

### Description One


### Plot Two
```{r echo=FALSE, Plot_Two}

```

### Description Two


### Plot Three
```{r echo=FALSE, Plot_Three}

```

### Description Three

------

# Reflection

> **Tip**: Here's the final step! Reflect on the exploration you performed and
the insights you found. What were some of the struggles that you went through?
What went well? What was surprising? Make sure you include an insight into
future work that could be done with the dataset.

> **Tip**: Don't forget to remove this, and the other **Tip** sections before
saving your final work and knitting the final report!