Red Wine Physicochemical Properties and Quality by Oliver Koch
==============================================================


```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using in your analysis in this code
# chunk.

# Notice that the parameter "echo" was set to FALSE for this code chunk. This
# prevents the code from displaying in the knitted HTML output. You should set
# echo=FALSE for all code chunks in your file, unless it makes sense for your
# report to show the code that generated a particular plot.

# The other parameters for "message" and "warning" should also be set to FALSE
# for other code chunks once you have verified that each plot comes out as you
# want it to. This will clean up the flow of your report.

library(readr)
library(tidyr)
library(knitr)
library(ggplot2)
library(dplyr)
library(GGally)
library(tibble)

```

```{r echo=FALSE, message=FALSE, warning=FALSE, Load_the_Data}
# Load the Data

# the first column of the csv file contains a row index without column name.
# R automatically assigns this column a name 'X1'

wine <- read_csv("wineQualityReds.csv", 
                      col_types = cols(quality = col_integer()))
wine <- reshape::rename(wine, c(X1 = "ID"))
```


## Introduction to Dataset


The dataset is related to red variants of the Portuguese "Vinho 
Verde" wine. It contains physical and chemical properties from objective 
measurements and a sensory assessment of the quality of the wine (median of at 
least 3 evaluations made by wine experts).

The dataset is available at this URL:
https://s3.amazonaws.com/udacity-hosted-downloads/ud651/wineQualityReds.csv

The details are described in 
_P. Cortez, A. Cerdeira, F. Almeida, T. Matos and J. Reis. 
Modeling wine preferences by data mining from physicochemical properties.
In Decision Support Systems, Elsevier, 47(4):547-553. ISSN: 0167-9236_, 
available at http://www3.dsi.uminho.pt/pcortez/wine5.pdf


### Attribute information:

Input variables (based on physicochemical tests):

1. Fixed acidity (tartaric acid - g / dm^3): most acids involved with wine or 
   fixed or nonvolatile (do not evaporate readily)
2. Volatile acidity (acetic acid - g / dm^3): the amount of acetic acid in wine, 
   which at too high of levels can lead to an unpleasant, vinegar taste
3. Citric acid (g / dm^3): found in small quantities, citric acid can add 
   'freshness' and flavor to wines
4. Residual sugar (g / dm^3): the amount of sugar remaining after fermentation 
   stops, it's rare to find wines with less than 1 gram/liter and wines with 
   greater than 45 grams/liter are considered sweet
5. Chlorides (sodium chloride - g / dm^3): the amount of salt in the wine
6. Free sulfur dioxide (mg / dm^3): the free form of SO2 exists in equilibrium 
   between molecular SO2 (as a dissolved gas) and bisulfite ion; it prevents 
   microbial growth and the oxidation of wine
7. Total sulfur dioxide (mg / dm^3): amount of free and bound forms of S02; in 
   low concentrations, SO2 is mostly undetectable in wine, but at free SO2 
   concentrations over 50 ppm, SO2 becomes evident in the nose and taste of wine
8. Density (g / cm^3): the density of water is close to that of water depending 
   on the percent alcohol and sugar content
9. pH: describes how acidic or basic a wine is on a scale from 0 (very acidic) 
   to 14 (very basic); most wines are between 3-4 on the pH scale
10. Sulphates (potassium sulphate - g / dm3): sulphates: a wine additive which 
    can contribute to sulfur dioxide gas (S02) levels, wich acts as an 
    antimicrobial and antioxidant
11. Alcohol (% by volume): the percent alcohol content of the wine

Output variable (based on sensory data): 

12. Quality. Score between 0 (very bad) and 10 (very good)



# Univariate Plots Section

An array of histograms and an array of boxplots provide a first overview of the 
ranges of values and the distribution of data:

```{r echo=FALSE, message=FALSE, warning=FALSE, histo_1}

wine %>%
gather(key="wine.att", value="measurement", -ID) %>%
ggplot() +
 geom_histogram(mapping=aes(measurement), bins=30) +
 facet_wrap(vars(wine.att), scales='free')
  
```
```{r echo=FALSE, message=FALSE, warning=FALSE, box_1}

wine %>%
select(-c(ID, quality)) %>%
gather(key="wine.att", value="measurement") %>%
ggplot(mapping=aes(measurement)) +
 geom_boxplot() +
 facet_wrap(vars(wine.att), scales='free')
  
```

At first glance, pH, density and quality have symmetric, normal-like 
distributions. The other variables are right skewed.

Plotting the histograms again with a logarithmic scale provides a better view on 
extreme values:

```{r echo=FALSE, message=FALSE, warning=FALSE, histo_2}
wine %>%
  gather(key="wine.att", value="measurement", -ID) %>%
  ggplot() +
   geom_histogram(mapping=aes(measurement)) +
   scale_x_log10() +
   facet_wrap(vars(wine.att), scales='free')
```

In order to better visualise the variables with long-tailed distributions, I 
create a filtered dataset (with manually set limits per variable) and plot the 
histograms for that:


```{r echo=FALSE, message=FALSE, warning=FALSE}
# dataset without extreme values
wine.typical <- filter(wine, alcohol < 13 & 
                              chlorides < 0.15 & 
                              citric.acid > 0.01 &
                              citric.acid < 0.8 & 
                              fixed.acidity < 12 &
                              free.sulfur.dioxide < 40 &
                              residual.sugar < 3.5 &
                              sulphates < 1 &
                              total.sulfur.dioxide < 100 &
                              volatile.acidity < 1
                           ) 

wine.typical %>%
gather(key="wine.att", value="measurement", -ID) %>%
filter(measurement > 0) %>% # cecause of log
ggplot() +
 geom_histogram(mapping=aes(measurement), bins=20) +
 scale_x_log10() +
 facet_wrap(vars(wine.att), scales='free')
```

Volatile acidity appears to have a bimodal distribution. Plotting this variable 
with a smaller bin width:

```{r echo=FALSE, message=FALSE, warning=FALSE}

wine %>%
ggplot() +
 geom_histogram(mapping=aes(volatile.acidity), binwidth=0.01) 
```
At this finer resolution, it's not so obvious that there are two modes.

Looking at quality:

```{r echo=FALSE, message=FALSE, warning=FALSE}

wine %>%
ggplot() +
 geom_histogram(mapping=aes(quality)) 

```

```{r echo=FALSE, message=FALSE, warning=FALSE}

table(wine$quality)
```

Quality can take values form 0 to 10. The wines present in the dataset only span 
a range from 3 to 8. There are no _very good_ or _very bad_ wines.

Introducting a new categorical variable, reducing quality to three levels:

- 0-3: bad
- 4-6: medium
- 7-10: good


```{r echo=FALSE, message=FALSE, warning=FALSE}
quality_groups <- ifelse(wine$quality <= 4, "bad", 
                         ifelse(wine$quality >= 6, "good", "medium"))
wine$bad_medium_good <- ordered(quality_groups, levels = c("bad", "medium", "good"))
table(wine$bad_medium_good)
```
There are only very few bad wines.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(wine) +
 geom_bar(mapping=aes(bad_medium_good)) 
```
Quality is distributed very unevenly - the majority of wines is in the "medium" 
group, and there are only very few bad wines.

# Univariate Analysis

Summary statistics of data:

```{r echo=FALSE, message=FALSE, warning=FALSE}

wine %>%
select(-ID) %>%
summary()
  
```

### What is the structure of your dataset?

```{r echo=FALSE, message=FALSE, warning=FALSE}
dim(wine)
```
The dataset has 12 variables and 1599 observations. All variables except quality
are real numbers. Quality is an integer variable.
There are no missing values. There are no obviously incorrect values (like 
negative values.)

### What is/are the main feature(s) of interest in your dataset?

The wine quality is the main feature of interest. As suggested in the datasets
documentation, it was created the purpose of developing regression models that
can predict the wine quality from physical & chemical measurement results.

### What other features in the dataset do you think will help support your investigation into your feature(s) of interest?

To me, the most interesting question is which variables have the strongest 
influence on wine quality. Potentially, all variables may have an influence.

### Did you create any new variables from existing variables in the dataset?

I created a categorical variable 'bad_medium_good' from quality. 

### Of the features you investigated, were there any unusual distributions? Did you perform any operations on the data to tidy, adjust, or change the form of the data? If so, why did you do this?

Alcohol and citric acid content have highly asymmetrical, almost triangluar 
distributions. 

Quality has a bell-shaped distribution, and there are few good ones and even 
fewer bad ones. 



# Bivariate Plots Section

A pair plot (scatterplot matrix) provides a first overview of dependencies and 
potential correlations between variables:

```{r echo=FALSE, message=FALSE, warning=FALSE}

wine %>%
   #sample_n(20) %>%
   select(-c(ID, bad_medium_good)) %>%
   ggpairs(upper = list(continuous = wrap("cor", size=2))) +
   theme_grey(base_size = 6)

   
```
Correlations between quality and (> 0.25) ...

- volatile acidity
- alcohol
- sulphates

also (> 0.5): 

- fixed.acidity: citric acid, density, pH
- volatile.acidity: citric acid
- citric acid: pH
- total.sulfur.dioxide: free.sulfur.dioxide

same, but with discrete target:

```{r echo=FALSE, message=FALSE, warning=FALSE}
wine %>%
  #sample_n(20) %>%
  select(-c(ID, quality)) %>%
  ggpairs(upper = list(continuous = wrap("cor", size=2)), mapping = aes(color = bad_medium_good)) +
  theme_grey(base_size = 6)
```




```{r echo=FALSE, message=FALSE, warning=FALSE}

wine.typical %>%
#sample_n(300) %>%
gather(key="wine.att", value="measurement", -c(ID, quality))%>%
ggplot(aes(x=measurement, y=quality)) +
 #geom_jitter(width=0, height=0.5) +
 geom_point(alpha=0.1) +
 # scale_x_sqrt() +
 #scale_x_log10() +
 geom_smooth() +
 #geom_quantile() +
 facet_wrap(vars(wine.att), scales='free')
```

#### related: pH, fixed.acidity, volatile.acidity, citric.acid

```{r echo=FALSE, message=FALSE, warning=FALSE}
wine.typical %>%
select(-c(residual.sugar, chlorides, free.sulfur.dioxide, total.sulfur.dioxide, density, sulphates, alcohol, quality)) %>%
gather(key="wine.att", value="measurement", -c(ID, pH)) %>%
ggplot(aes(x=measurement, y=pH)) +
 #geom_jitter(width=0, height=0.5) +
 geom_point(alpha=0.1) +
 # scale_x_sqrt() +
 #scale_x_log10() +
 geom_smooth() +
 #geom_quantile() +
 facet_wrap(vars(wine.att), scales='free')
```

- dependency between pH and fixed.acidity
- no clear dependency on volatile.acidity

#### related: free.sulfur.dioxide, total.sulfur.dioxide

```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(data=wine, aes(x=free.sulfur.dioxide, y=total.sulfur.dioxide)) +
   geom_point() +
   geom_abline(slope=1, intercept=0)

```
```{r echo=FALSE, message=FALSE, warning=FALSE}
bound.sulfur.dioxide <- wine$total.sulfur.dioxide - wine$free.sulfur.dioxide
cor(wine$total.sulfur.dioxide, wine$free.sulfur.dioxide)
cor(bound.sulfur.dioxide, wine$free.sulfur.dioxide)
```


```{r echo=FALSE, message=FALSE, warning=FALSE}

wine.typical %>%
   mutate(bound.sulfur.dioxide = total.sulfur.dioxide - free.sulfur.dioxide) %>%
   ggplot(aes(x = bound.sulfur.dioxide, y=free.sulfur.dioxide)) +
   geom_point() +
   geom_smooth()

wine.typical %>%
   mutate(bound.sulfur.dioxide = total.sulfur.dioxide - free.sulfur.dioxide) %>%
   ggplot(aes(x = bound.sulfur.dioxide, y=free.sulfur.dioxide)) +
   geom_point() +
   #scale_x_log10() +
   scale_y_log10() +
   geom_smooth()

```

- all points above line
- could introduce a new variable: bound = total - free

Intro says that for free SO2 > 50 ppm quality suffers?
Units are mg/dm^3 -> that is 

```{r echo=FALSE, message=FALSE, warning=FALSE}


wine %>%
#select(c(free.sulfur.dioxide, quality)) %>%
#gather(key="wine.att", value="measurement") %>%
ggplot(mapping=aes(free.sulfur.dioxide)) +
 geom_boxplot() +
 facet_grid(rows = vars(quality), scales='free')
  
```


```{r echo=FALSE, message=FALSE, warning=FALSE}

wine %>%
   filter(free.sulfur.dioxide >= 50) %>%
   select(quality) %>%
   table
```

- wines with free.sulfur.dioxide > 50 are in medium to good range
- wines of quality 5 have the highest median free.sulfur.dioxide
- with increasing quality, median free.sulfur.dioxide decreases
- with decreasing quality, median free.sulfur.dioxide decreases
- statement from dataset description cannot be observed in this dataset.


```{r echo=FALSE, message=FALSE, warning=FALSE}


wine %>%
#select(c(free.sulfur.dioxide, quality)) %>%
#gather(key="wine.att", value="measurement") %>%
ggplot(mapping=aes(sulphates)) +
 geom_boxplot() +
 facet_grid(rows = vars(bad_medium_good), scales='free')
  
```



```{r echo=FALSE, message=FALSE, warning=FALSE}


wine %>%
#select(c(free.sulfur.dioxide, quality)) %>%
#gather(key="wine.att", value="measurement") %>%
ggplot(mapping=aes(sulphates)) +
 geom_boxplot() +
 facet_grid(rows = vars(quality), scales='free')
  
```

- good wines tend to have higher sulphate content (median 0.7), but not too high
- wines with sulphate content > 1.4 don't make it into the upper range

```{r echo=FALSE, message=FALSE, warning=FALSE}
wine %>%
   select(c(sulphates, quality)) %>%
   filter(quality >= 7) %>%
   summary()
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
wine %>%
   select(c(sulphates, quality)) %>%
   filter(quality <= 4) %>%
   summary()
```

#### related: density, alcohol, sugar


```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(data=wine, aes(x=residual.sugar, y=density)) + 
   geom_point() +
   #scale_x_log10() +
   coord_cartesian(xlim = c(0, 4)) +
   geom_smooth() 

ggplot(data=wine, aes(x=alcohol, y=density)) + 
   geom_point() +
   geom_smooth() 
```

build a linear model: density depending on 


# Bivariate Analysis

> **Tip**: As before, summarize what you found in your bivariate explorations
here. Use the questions below to guide your discussion.

### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?

### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?

### What was the strongest relationship you found?

- alcohol / sugar / acidity -> quality
- alcohol / sugar -> density
- there is some redundancy here: density depends on alcohol, sugar and other
  stuff

# Multivariate Plots Section

> **Tip**: Now it's time to put everything together. Based on what you found in
the bivariate plots section, create a few multivariate plots to investigate
more complex interactions between variables. Make sure that the plots that you
create here are justified by the plots you explored in the previous section. If
you plan on creating any mathematical models, this is the section where you
will do that.

```{r echo=FALSE, message=FALSE, warning=FALSE, Multivariate_Plots}

```

#### sulfates / alcohol / quality
```{r echo=FALSE, message=FALSE, warning=FALSE}
#saq <- select(wine.filtered, c(volatile.acidity, alcohol, quality))

#ggplot(data=saq) +
#   geom_bin2d(aes(x=volatile.acidity, y=alcohol))
   #geom_point() +
   #geom_contour(aes(z=quality))

qlevel <- ordered(wine$quality, levels = 0:10)
wine %>%
  add_column(qlevel = qlevel) %>%
  ggplot() + 
   # geom_point(mapping=aes(x=alcohol, y=volatile.acidity, color=qlevel), alpha=0.9) +
   geom_jitter(mapping=aes(x=alcohol, y=volatile.acidity, color=qlevel), alpha=0.9, width=0.005, height=0) +
   #geom_contour(aes(x=alcohol, y=volatile.acidity, z=qlevel), )
   #scale_x_log10() +
   scale_y_sqrt()


```

```{r echo=FALSE, message=FALSE, warning=FALSE}

wine %>%
  #add_column(qlevel = qlevel) %>%
  ggplot() + 
   # geom_point(mapping=aes(x=alcohol, y=volatile.acidity, color=qlevel), alpha=0.9) +
   geom_jitter(mapping=aes(x=chlorides, y=sulphates, color=bad_medium_good), alpha=0.9, width=0.005, height=0) +
   #geom_contour(aes(x=alcohol, y=volatile.acidity, z=qlevel), )
   scale_x_log10() + 
   scale_y_log10()


```


```{r echo=FALSE, message=FALSE, warning=FALSE}

wine %>%
  #add_column(qlevel = qlevel) %>%
  ggplot() + 
   # geom_point(mapping=aes(x=alcohol, y=volatile.acidity, color=qlevel), alpha=0.9) +
   geom_jitter(mapping=aes(x=alcohol, y=sulphates, color=bad_medium_good), alpha=0.9, width=0.005, height=0) +
   #geom_contour(aes(x=alcohol, y=volatile.acidity, z=qlevel), )
   #scale_x_log10() +
   scale_y_sqrt()


```


- good wines tend to have high alcohol and low volatile acidity

```{r echo=FALSE, message=FALSE, warning=FALSE}
#qlevel <- ordered(wine$quality, levels = 0:10)
wine %>%
  #add_column(qlevel = qlevel) %>%
  ggplot() + 
   # geom_point(mapping=aes(x=alcohol, y=volatile.acidity, color=qlevel), alpha=0.9) +
   geom_jitter(mapping=aes(x=alcohol, y=sulphates, color=bad_medium_good), alpha=0.9, width=0.005, height=0) +
   #geom_contour(aes(x=alcohol, y=volatile.acidity, z=qlevel), )
   #scale_x_log10() +
   scale_y_log10()

```



```{r echo=FALSE, message=FALSE, warning=FALSE}
wine %>%
   ggplot() +
   geom_point(mapping=aes(x=alcohol, y=volatile.acidity, color=bad_medium_good), alpha=0.9, width=0.005, height=0)

```


```

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?

### Were there any interesting or surprising interactions between features?

### OPTIONAL: Did you create any models with your dataset? Discuss the \
strengths and limitations of your model.





------

# Final Plots and Summary

> **Tip**: You've done a lot of exploration and have built up an understanding
of the structure of and relationships between the variables in your dataset.
Here, you will select three plots from all of your previous exploration to
present here as a summary of some of your most interesting findings. Make sure
that you have refined your selected plots for good titling, axis labels (with
units), and good aesthetic choices (e.g. color, transparency). After each plot,
make sure you justify why you chose each plot by describing what it shows.

### Plot One
```{r echo=FALSE, Plot_One}

```

### Description One


### Plot Two
```{r echo=FALSE, Plot_Two}

```

### Description Two


### Plot Three
```{r echo=FALSE, Plot_Three}

```

### Description Three

------

# Reflection

> **Tip**: Here's the final step! Reflect on the exploration you performed and
the insights you found. What were some of the struggles that you went through?
What went well? What was surprising? Make sure you include an insight into
future work that could be done with the dataset.

> **Tip**: Don't forget to remove this, and the other **Tip** sections before
saving your final work and knitting the final report!