Red Wine Physicochemical Properties and Quality by Oliver Koch
==============================================================


```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using in your analysis in this code
# chunk.

# Notice that the parameter "echo" was set to FALSE for this code chunk. This
# prevents the code from displaying in the knitted HTML output. You should set
# echo=FALSE for all code chunks in your file, unless it makes sense for your
# report to show the code that generated a particular plot.

# The other parameters for "message" and "warning" should also be set to FALSE
# for other code chunks once you have verified that each plot comes out as you
# want it to. This will clean up the flow of your report.

library(readr)
library(tidyr)
library(knitr)
library(ggplot2)
library(dplyr)
library(GGally)
library(tibble)
```

```{r echo=FALSE, message=FALSE, warning=FALSE, Load_the_Data}
# Load the Data
# the first column of the csv file contains a row index without column name.
# R automatically assigns this column a name 'X1'

wine <- read_csv("wineQualityReds.csv", 
                      col_types = cols(quality = col_integer()))
wine <- reshape::rename(wine, c(X1 = "ID"))
```


## Introduction to the Dataset

The dataset is related to red variants of the Portuguese "Vinho 
Verde" wine. It contains physical and chemical properties from objective 
measurements and a sensory assessment of the quality of the wine (median of at 
least 3 evaluations made by wine experts).

The dataset is available at this URL:
https://s3.amazonaws.com/udacity-hosted-downloads/ud651/wineQualityReds.csv

The details are described in 
_P. Cortez, A. Cerdeira, F. Almeida, T. Matos and J. Reis. 
Modeling wine preferences by data mining from physicochemical properties.
In Decision Support Systems, Elsevier, 47(4):547-553. ISSN: 0167-9236_, 
available at http://www3.dsi.uminho.pt/pcortez/wine5.pdf


### Attribute information:

Input variables (based on physicochemical tests):

1. Fixed acidity (tartaric acid - g / dm^3): most acids involved with wine or 
   fixed or nonvolatile (do not evaporate readily)
2. Volatile acidity (acetic acid - g / dm^3): the amount of acetic acid in wine, 
   which at too high of levels can lead to an unpleasant, vinegar taste
3. Citric acid (g / dm^3): found in small quantities, citric acid can add 
   'freshness' and flavor to wines
4. Residual sugar (g / dm^3): the amount of sugar remaining after fermentation 
   stops, it's rare to find wines with less than 1 gram/liter and wines with 
   greater than 45 grams/liter are considered sweet
5. Chlorides (sodium chloride - g / dm^3): the amount of salt in the wine
6. Free sulfur dioxide (mg / dm^3): the free form of SO2 exists in equilibrium 
   between molecular SO2 (as a dissolved gas) and bisulfite ion; it prevents 
   microbial growth and the oxidation of wine
7. Total sulfur dioxide (mg / dm^3): amount of free and bound forms of S02; in 
   low concentrations, SO2 is mostly undetectable in wine, but at free SO2 
   concentrations over 50 ppm, SO2 becomes evident in the nose and taste of wine
8. Density (g / cm^3): the density of water is close to that of water depending 
   on the percent alcohol and sugar content
9. pH: describes how acidic or basic a wine is on a scale from 0 (very acidic) 
   to 14 (very basic); most wines are between 3-4 on the pH scale
10. Sulphates (potassium sulphate - g / dm3): sulphates: a wine additive which 
    can contribute to sulfur dioxide gas (S02) levels, wich acts as an 
    antimicrobial and antioxidant
11. Alcohol (% by volume): the percent alcohol content of the wine

Output variable (based on sensory data): 

12. Quality. Score between 0 (very bad) and 10 (very good)


# Univariate Plots Section

An array of histograms and an array of boxplots provide a first overview of the 
ranges of values and the distribution of data:

```{r echo=FALSE, message=FALSE, warning=FALSE, histo_1}
wine %>%
gather(key="wine.att", value="measurement", -ID) %>%
ggplot() +
 geom_histogram(mapping=aes(measurement), bins=50) +
 facet_wrap(vars(wine.att), scales='free')
```
```{r echo=FALSE, message=FALSE, warning=FALSE, box_1}
wine %>%
select(-c(ID, quality)) %>%
gather(key="wine.att", value="measurement") %>%
ggplot(mapping=aes(measurement)) +
 geom_boxplot() +
 facet_wrap(vars(wine.att), scales='free')
```

At a first glance, pH, density and quality have symmetric, normal-like 
distributions. The other variables are right skewed.

Plotting the histograms again with a logarithmic scale provides a better view on 
extreme values:

```{r echo=FALSE, message=FALSE, warning=FALSE, histo_2}
wine %>%
  gather(key="wine.att", value="measurement", -ID) %>%
  ggplot() +
   geom_histogram(mapping=aes(measurement), bins=50) +
   scale_x_log10() +
   facet_wrap(vars(wine.att), scales='free')
```

In order to better visualize the variables with long-tailed distributions, I 
create a filtered dataset (with manually set limits per variable) and plot the 
histograms again:

```{r echo=FALSE, message=FALSE, warning=FALSE}
# dataset without extreme values
wine.typical <- filter(wine, alcohol < 13 & 
                              chlorides < 0.15 & 
                              citric.acid > 0.01 &
                              citric.acid < 0.8 & 
                              fixed.acidity < 12 &
                              free.sulfur.dioxide < 40 &
                              residual.sugar < 3.5 &
                              sulphates < 1 &
                              total.sulfur.dioxide < 100 &
                              volatile.acidity < 1
                           ) 

wine.typical %>%
gather(key="wine.att", value="measurement", -ID) %>%
filter(measurement > 0) %>% # because of log
ggplot() +
 geom_histogram(mapping=aes(measurement), bins=20) +
 scale_x_log10() +
 facet_wrap(vars(wine.att), scales='free')
```

Looking at quality:

```{r echo=FALSE, message=FALSE, warning=FALSE}
wine %>%
ggplot() +
 geom_histogram(mapping=aes(quality)) 
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
table(wine$quality)
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
summary(wine$quality)
```

Quality can take values form 0 to 10. The wines present in the dataset only span 
a range from 3 to 8. There are no _very good_ or _very bad_ wines; the majority 
has a quality of 5 or 6.

Introducing a new categorical variable, reducing quality to three grades:

- 0-3: bad
- 4-6: medium
- 7-10: good

```{r echo=FALSE, message=FALSE, warning=FALSE}
quality_grades <- ifelse(wine$quality <= 4, "bad", 
                         ifelse(wine$quality >= 7, "good", "medium"))
wine$grade <- ordered(quality_grades, levels = c("bad", "medium", "good"))
table(wine$grade)
```

There are only very few bad wines.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(wine) +
 geom_bar(mapping=aes(grade)) 
```

Quality is distributed very unevenly - the majority of wines is in the "medium" 
group, and there are only very few bad wines.


# Univariate Analysis

Summary statistics of data:

```{r echo=FALSE, message=FALSE, warning=FALSE}

wine %>%
select(-ID) %>%
summary()
  
```

### What is the structure of your dataset?

```{r echo=FALSE, message=FALSE, warning=FALSE}
dim(wine)
```

The dataset has 12 variables and 1599 observations. All variables except quality
are real numbers. Quality is an integer variable.
There are no missing values. There are no obviously incorrect values (like 
negative values.)


### What is/are the main feature(s) of interest in your dataset?

The wine quality is the main feature of interest. As suggested in the datasets
documentation, it was created the purpose of developing regression models that
can predict the wine quality from physical & chemical measurement results.


### What other features in the dataset do you think will help support your investigation into your feature(s) of interest?

To me, the most interesting question is which variables have the strongest 
influence on wine quality. Potentially, all variables may have an influence.


### Did you create any new variables from existing variables in the dataset?

I created a categorical variable 'grade' from quality. 


### Of the features you investigated, were there any unusual distributions? Did you perform any operations on the data to tidy, adjust, or change the form of the data? If so, why did you do this?

Alcohol and citric acid content have highly asymmetrical, almost triangular 
distributions. 

Quality has a bell-shaped distribution, there are few good ones and even 
fewer bad ones. 


# Bivariate Plots Section

A pair plot (scatterplot matrix) provides a first overview of dependencies and 
potential correlations between variables:

```{r echo=FALSE, message=FALSE, warning=FALSE}
wine %>%
   select(-c(ID, grade)) %>%
   ggpairs(upper = list(continuous = wrap("cor", size=2))) +
   theme_grey(base_size = 6)
```

Again a pair plot, this time using the categorical quality value:

```{r echo=FALSE, message=FALSE, warning=FALSE}
wine %>%
  select(-c(ID, quality)) %>%
  ggpairs(upper = list(continuous = wrap("cor", size=2)), 
          mapping = aes(color = grade)) +
  theme_grey(base_size = 6)
```

For the main variable of interest (quality), the highest correlations 
are with volatile acidity, alcohol, and sulphates. 

High values of correlation also occur in these groups: 

- volatile acidity, fixed.acidity, citric acid, density, pH
- density, alcohol, pH, fixed.acidity, citric.acid, residual.sugar


### Which variables influence quality?

The following plots highlight patterns between the measured variables and the 
wine quality:

```{r echo=FALSE, message=FALSE, warning=FALSE}
wine %>%
gather(key="wine.att", value="measurement", -c(ID, quality, grade)) %>%
ggplot(aes(x=measurement, y=quality)) +
 geom_point(alpha=0.1) +
 geom_smooth() +
 facet_wrap(vars(wine.att), scales='free')
```

- alcohol, density, total.sulfur.dioxide, volatile acidity
- free sulfur dioxide and residual sugar do not have an influence. 
- nonlinear relationships (which are not captured by linear correlation measures 
  or linear models): citric acid, chlorides, sulphates


### Which variables influence the pH value?

```{r echo=FALSE, message=FALSE, warning=FALSE}
wine %>%
select(-c(residual.sugar, chlorides, free.sulfur.dioxide, total.sulfur.dioxide, 
          density, sulphates, alcohol, quality, grade)) %>%
gather(key="wine.att", value="measurement", -c(ID, pH)) %>%
ggplot(aes(x=measurement, y=pH)) +
 geom_point(alpha=0.1) +
 geom_smooth() +
 facet_wrap(vars(wine.att), scales='free')
```

correlation between pH and fixed.acidity:

```{r echo=FALSE, message=FALSE, warning=FALSE}
cor(wine$fixed.acidity, wine$pH)
```

correlation between pH and volatile acidity:

```{r echo=FALSE, message=FALSE, warning=FALSE}
cor(wine$volatile.acidity, wine$pH)
```

correlation between pH and citric acid:

```{r echo=FALSE, message=FALSE, warning=FALSE}
cor(wine$citric.acid, wine$pH)
```

Conclusion:

- medium correlation between pH and citric.acid and fixed.acidity
- no clear correlation between pH and volatile.acidity



# Bivariate Analysis

### Talk about some of the relationships you observed in this part of the investigation. How did the feature(s) of interest vary with other features in the dataset?

The alcohol content has the strongest influence on wine quality. The variables
with the next highest correlation coefficients are sulphates and volatile 
acidity. 


### Did you observe any interesting relationships between the other features (not the main feature(s) of interest)?

Relationships exist 

- among acidity related variables (pH, citric acid, fixed acidity, volatile 
  acidity). 
- between density and almost all other variables (because the density of the 
  wine depends on the concentration of the ingredients). 
  Most prominent influences from alcohol and residual sugar.


### What was the strongest relationship you found?

The strongest (linear) relationship is between total sulfur dioxide and free 
sulfur dioxide.


# Multivariate Plots Section

### Explore relationships between quality, alcohol, volatile acidity and sulphates

```{r echo=FALSE, message=FALSE, warning=FALSE, Multivariate_Plots}
wine %>%
  ggplot() + 
   geom_point(mapping=aes(x=alcohol, y=volatile.acidity, color=grade), 
              alpha=0.9) 
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
wine %>%
  ggplot() + 
   geom_point(mapping=aes(x=alcohol, y=sulphates, color=grade), 
              alpha=0.9) 
```


```{r echo=FALSE, message=FALSE, warning=FALSE}

wine %>%
  ggplot() + 
   geom_point(mapping=aes(x=volatile.acidity, y=sulphates, color=grade), 
              alpha=0.8) 

```

In the plots above, bad and good wines are separated - if there were just good 
ones and bad ones, the three variables could be used to classify wines.
However, the majority of wines are in the medium group, and they cover the 
whole range of values, making a classification (with the given variables and 
linear decision boundaries) difficult.


# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the investigation. Were there features that strengthened each other in terms of looking at your feature(s) of interest?

In the plots above, bad and good wines are separated - if there were just good 
ones and bad ones, the three variables could be used to classify wines.
However, the majority of wines are in the medium group, and they cover the 
whole range of values, making a classification (with the given variables and 
linear decision boundaries) difficult.

------

# Final Plots and Summary


### Plot One

```{r echo=FALSE, Plot_One}
ggplot(data = wine) +
   geom_bar(aes(quality)) +
   labs(x = "Quality", title = "Distribution of wine quality")
```

### Description One

On a possible range from 0 to 10, all wines fall 
between 3 and 8, and the majority has quality ratings of 5 or 6.


### Plot Two

```{r echo=FALSE, message=FALSE, warning=FALSE, Plot_Two}

wine %>%
  select(c(fixed.acidity, volatile.acidity, citric.acid, sulphates, alcohol, 
           grade)) %>%
   ggpairs(upper = list(continuous = wrap("cor", size=2.5)), 
           mapping = aes(color = grade)) +
     theme_grey(base_size = 8) +
     labs(title = "Scatterplot matrix for selected variables")
```

### Description Two

The scatterplot matrix shows those variable where distributions differ among  
wine "grades". This subset of variables could be a starting point for building 
classification or regression models.


### Plot Three

```{r echo=FALSE, message=FALSE, warning=FALSE, Plot_Three}
wine %>%
gather(key="wine.att", value="Attribute", 
       -c(ID, quality, grade, fixed.acidity, free.sulfur.dioxide, residual.sugar, 
          pH, density)) %>%
ggplot(aes(x=Attribute, y=quality)) +
 geom_point(alpha=0.1) +
 geom_smooth() +
 facet_wrap(vars(wine.att), scales='free') +
 labs(title = "Linear and nonlinear relationships between quality and other attributes", 
      y = "Quality")

```

### Description Three

This plot shows that there are nonlinear relationships between quality and some
of the other variables, e.g. citric acid. A good classification or regression 
model would have to capture these nonlinear relationships.

------

# Reflection

> **Tip**: Here's the final step! Reflect on the exploration you performed and
the insights you found. What were some of the struggles that you went through?
What went well? What was surprising? Make sure you include an insight into
future work that could be done with the dataset.

> **Tip**: Don't forget to remove this, and the other **Tip** sections before
saving your final work and knitting the final report!